---
title: "EFSA Submission v1"
subtitle: "Data export of trace elements for EFDA submission"
---

```{css, echo=FALSE}
th {
  white-space: nowrap;
}
```

The page presents the data export of trace elements in background sediments from the pilot database.

```{r}
library(dplyr)
library(tibble)

efsa_submission <- tribble(
  ~Column,            ~`Lookup`, ~Optionality, ~Table,             ~`Column Names`,           ~`Fixed Text`,      ~Missing,
  "sampleId",         NA,        "Mandatory",  "Sample",           "sample_id",               NA,                 NA,
  "recId",            "Yes",     "Mandatory",  NA,                 NA,                        NA,                 NA,
  "envComp",          NA,        "Mandatory",  NA,                 NA,                        "marine sediment",  NA,
  "pristineLoc",      NA,        "Mandatory",  NA,                 NA,                        "Yes",              NA,
  "sampleLocGC",      NA,        "Mandatory",  "Site",             "lat + lon",               NA,                 NA,
  "sampleLocCM",      NA,        "Mandatory",  "Site",             "country + municipality",  NA,                 NA,
  "sampleDate",       NA,        "Mandatory",  "Sample",           "sample_time",             NA,                 NA,
  "traceElText",      NA,        "Mandatory",  "Parameter",        "param_name",              NA,                 NA,
  "traceElCode",      "Yes",     "Mandatory",  NA,                 NA,                        NA,                 NA,
  "specText",         NA,        "Mandatory",  "Parameter",        'param_name + " (Total)"', NA,                 NA,
  "specCode",         "Yes",     "Mandatory",  NA,                 NA,                        NA,                 NA,
  "conc",             NA,        "Mandatory",  "Sediment",         "value",                   NA,                 NA,
  "unitText",         NA,        "Mandatory",  "Analysis Method",  "unit",                    NA,                 NA,
  "unitCode",         "Yes",     "Mandatory",  NA,                 NA,                        NA,                 NA,
  "weightText",       NA,        "Mandatory",  NA,                 NA,                        "dw",               NA,
  "weightCode",       "Yes",     "Mandatory",  NA,                 NA,                        NA,                 NA,
  "methAnText",       NA,        "Mandatory",  "Analysis Method",  "analysis",                NA,                 NA,
  "methAnCode",       "Yes",     "Mandatory",  NA,                 NA,                        NA,                 NA,
  "LOD",              NA,        "Optional",   "LLD",              "value (type = 'LOD')",    NA,                 NA,
  "LOQ",              NA,        "Optional",   "LLD",              "value (type = 'LOQ')",    NA,                 NA,
  "accLab",           NA,        "Optional",   NA,                 NA,                        "No",               NA,
  "phSed",            NA,        "Optional",   NA,                 NA,                        NA,                 "Yes",
  "TextureSedClay",   NA,        "Optional",   "Sediment",         "value",                   NA,                 NA,
  "TextureSedSilt",   NA,        "Optional",   "Sediment",         "value",                   NA,                 NA,
  "TextureSedSand",   NA,        "Optional",   "Sediment",         "value",                   NA,                 NA,
  "unitText2",        NA,        "Optional",   NA,                 NA,                        "%",                NA,
  "unitCode2",        "Yes",     "Optional",   NA,                 NA,                        NA,                 NA,
  "weightText2",      "Yes",     "Optional",   NA,                 NA,                        NA,                 NA,
  "weightCode2",      "Yes",     "Optional",   NA,                 NA,                        NA,                 NA,
  "ocSed",            NA,        "Optional",   "Sediment",         "value",                   NA,                 NA,
  "unitText3",        NA,        "Optional",   "Analysis Method",  "unit",                    NA,                 NA,
  "unitCode3",        "Yes",     "Optional",   NA,                 NA,                        NA,                 NA,
  "weightText3",      "Yes",     "Optional",   NA,                 NA,                        NA,                 NA,
  "weightCode3",      "Yes",     "Optional",   NA,                 NA,                        NA,                 NA,
  "phWater",          NA,        "Optional",   NA,                 NA,                        NA,                 "Yes",
  "hardWater",        NA,        "Optional",   NA,                 NA,                        NA,                 "Yes",
  "DOC",              NA,        "Optional",   NA,                 NA,                        NA,                 "Yes",
  "unitText5",        NA,        "Optional",   NA,                 NA,                        NA,                 NA,
  "unitCode5",        "Yes",     "Optional",   NA,                 NA,                        NA,                 NA,
  "publicData",       NA,        "Optional",   NA,                 NA,                        "Yes",              NA,
  "refPublication",   NA,        "Optional",   NA,                 NA,                        NA,                 "Yes",
  "comments",         NA,        "Optional",   NA,                 NA,                        NA,                 "Yes",
  "Missing recId",        "Yes", "Optional",   NA,                 NA,                        NA,                 NA,
  "Missing envComp",      "Yes", "Optional",   NA,                 NA,                        NA,                 NA,
  "Missing pristineLoc",  "Yes", "Optional",   NA,                 NA,                        NA,                 NA,
  "Missing sampleLoc",    "Yes", "Optional",   NA,                 NA,                        NA,                 NA,
  "Missing sampleDate",   "Yes", "Optional",   NA,                 NA,                        NA,                 NA,
  "Missing traceEl",      "Yes", "Optional",   NA,                 NA,                        NA,                 NA,
  "Missing spec",         "Yes", "Optional",   NA,                 NA,                        NA,                 NA,
  "Missing conc",         "Yes", "Optional",   NA,                 NA,                        NA,                 NA,
  "Missing methAn",       "Yes", "Optional",   NA,                 NA,                        NA,                 NA
)

efsa_submission_full <- efsa_submission %>%
  mutate(
    across(
      where(is.character),
      ~coalesce(.x, "")
    )
  ) %>%
  rename(c(`Source Table` = "Table",
           `Source Columns` = "Column Names"))

efsa_submission <- efsa_submission_full %>%
  filter(Lookup != "Yes") %>%
  dplyr::select(-c(Lookup, Missing))

```

## Data Mapping for Submission
```{r}
efsa_submission
```

## Full Data Mapping
```{r}
efsa_submission_full
```

## Data Export Method
As the EFSA submission tool is an Excel file containing embedded macros, the data must be mapped column by column rather than through a standard row‑based export. This approach ensures that the structure of the submission file is not disrupted.
