---
title: "Vannmiljø Pilot Database"
subtitle: "A pilot study to evaluate the database structure using the Vannmiljø data"
---

This site presents the initial version of the database design and its potential applications using the publicly available [Vannmiljø](https://vannmiljo.miljodirektoratet.no) dataset.

## Source Data

- Data search page at Vannmiljø: [Søk i vannrelaterte data](https://vannmiljo.miljodirektoratet.no/#/searchregistrations)

### Search criteria

#### Medium

 * Sediment saltvann

#### Parameter

 * Kobber
 * Sink
 * Mangan
 * Kobolt
 * Molybden
 * Selen

### Count of Sampling Sites
The table below summarizes the number of sampling sites documented across 27 activities and projects between 1975 and 2025.

```{r}
library(tidyverse)
library(DT)

# 1. Define the URL for the data file
data_file_url <- "https://github.com/seafood-hazards/vannmiljo-pilot/releases/download/v0.1.1/pilot_vannmiljo.tsv.gz"
local_data_file_name <- "pilot_vannmiljo.tsv.gz"

# 2. Download Logic
if (!file.exists(local_data_file_name)) {
  options(timeout = 600)
  download.file(data_file_url, destfile = local_data_file_name, mode = "wb")
  message("Database downloaded.")
} else {
  message("Using existing local database.")
}

df_vannmiljo_sediment <- read_tsv(local_data_file_name)
df_counts <- df_vannmiljo_sediment %>%
  distinct(activity_id, activity_name, site_code, site_name, sample_time, param_name) %>%
  mutate(year = year(sample_time)) %>%
  count(activity_id, activity_name, year, param_name) %>%
  pivot_wider(names_from = param_name, values_from = n) %>%
  replace(is.na(.), 0) %>%
  mutate(activity_id = factor(activity_id),
         activity_name = factor(activity_name)) %>%
  rename(`Activity ID` = activity_id,
         `Activity Name` = activity_name,
         Year = year)

datatable(df_counts, filter = "top", rownames = FALSE, class = "nowrap", options = list(scrollX = TRUE))

```

## Activities Covered by Vannmiljø

The Vannmiljø dataset contains sediment data for six chemical parameters from the following 27 activities and projects:

```{r}
library(readr)
library(dplyr)

read_tsv("./.data/vannmiljo_activities.tsv.gz") %>%
  mutate(across(where(is.numeric), ~ replace(., is.na(.), 0)))
```

## Database Design and Creation

After creating the ER (Entity–Relationship) diagram, the tables were implemented in a single SQLite file containing 10 tables. All required data were extracted from the Vannmiljø export file and imported into the corresponding tables.

- See the [DB Schema](./db-schema.qmd) page for details on the database architecture.
- See the [Data Preparation](./data-preparation.qmd) page for the data preparation process used prior to importing the Vannmiljø dataset.

## Geospatial Analysis

To facilitate efficient core and sample selection, distances to the nearest coastline were calculated for the Vannmiljø data. In addition, the nearest country and municipality, as well as the sea and ocean names, were identified based on the core locations.

- See the [Distance Calculation](./distance-to-coast.qmd) page for the methods and results of calculating the distance to the nearest coastline.
- See the [Location Names](./location-names.qmd) page for the estimation of location names from geographic coordinates.
- See the [Interactive Map](./distance-interactive-map.qmd) page for an interactive visualization of core locations with meta-information.

## Data Export

Data can be exported for both exploration and submission purposes.

- See the [Export to Tabular File](./data-export.qmd) page for a description of a single tabular file format (`.tsv.gz`) that is extracted and merged from all database tables for data exploration.

## Tools

- See the [Pilot DB Viewer](./pilot-db-viewer.qmd) page for a read-only view of the pilot database.
