---
title: "Sediment Particle Size Fractions"
subtitle: "From raw grain-size measurements to standardised fractions"
---

```{r}
library(tidyverse)
library(knitr)
library(kableExtra)
```

## Overview

Marine sediment cores collected as part of this project were analysed for grain-size
distribution. The raw data report particle abundance as percentages across a variety
of size classes and aggregate measures.

```{r}
tibble(
  Fraction = c("Clay", "Silt", "Sand", "Gravel"),
  `Size range` = c("< 2 µm", "2 – 63 µm", "63 – 2000 µm", "> 2000 µm"),
  Description = c(
    "Very fine particles, high surface area",
    "Fine particles, settles slowly",
    "Medium particles comprising five sub-fractions (63–125, 125–250, 250–500, 500–1000, 1000–2000 µm)",
    "Coarse particles > 2 mm; typically minor in open marine settings"
  )
) |>
  kable(align = c("l", "l", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

---

## Data Quality Issues

### 1. Non-standard measurement operators

Not all measurements are simple point estimates. Each value in the source data is
accompanied by an operator describing its relationship to the true value:

```{r}
tibble(
  Operator = c("=", "<", ">"),
  Meaning = c(
    "Exact measurement",
    "Below detection or quantification limit",
    "Exceeds the upper measurement range (lower bound only)"
  ),
  `Treatment` = c(
    "Used as-is",
    "Replaced with half the reported value (mid-point convention)",
    "Used as-is (treated as a measured lower bound)"
  )
) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

### 2. Decimal-point errors

Values well above 100 % were present in the source data, most likely caused by decimal
points being omitted or misplaced during data entry in spreadsheet software. For
example, a true value of 45.9 % was recorded as 45903.

### 3. Overlapping and redundant parameters

The source data contain both aggregate parameters and their constituent sub-fractions
for the same sample:

- **Fines < 63 µm** (`FINS`) is the sum of Clay + Silt, but some samples report
  FINS alongside separate Clay and Silt measurements, double-counting that portion.
- **Coarse > 63 µm** (`GSMF_63`) is similarly redundant when individual sand
  sub-fractions or a gravel measurement are also present.

Including both aggregate and component values inflates the apparent total well beyond
100 % and prevents straightforward fraction conversion.

---

## Processing Pipeline

The data are processed in six sequential steps. Each step feeds directly into the next.

### Step 0: Operator adjustment

Measurement operators are resolved into usable numeric values using the conventions
shown in the table above. All subsequent steps operate on these adjusted values.

### Step 0b: Decimal-point correction

For each parameter, a background median is computed from all valid values (i.e. those
already in the range 0–100 %). Values exceeding 100 % are then corrected by dividing
by successive powers of ten (10, 100, 1000, …) and selecting the candidate that is
both ≤ 100 % and closest to that parameter's background median. Values that cannot be
recovered this way are set to missing and flagged.

### Step 0c: Overlap removal and rescaling

Redundant aggregate parameters are removed from any sample where their components are
also present. Specifically:

- `FINS` is dropped if either `GSMF2` (Clay) or `GSMF2_63` (Silt) is present.
- `GSMF_63` is dropped if any sand sub-fraction or `GSMF_2000` (Gravel) is present.

After deduplication, the remaining values for each sample are summed. If the total
exceeds 100 %, all values are rescaled proportionally so that the sample sums to
exactly 100 %.

### Step 1: Wide-format restructuring

The cleaned long-format data are pivoted to wide format, giving one row per
sample × sediment layer with columns for each grain-size parameter and its operator.

### Step 2: Arithmetic derivation (four passes)

The four target fractions are derived algebraically from whichever measurements are
available, without using any background statistics. The derivation exploits the
following constraints.

 * **Clay** + **Silt** = **Fines**
 * **Sand** + **Gravel** = **Coarse**
 * **Fines** + **Coarse** = *100%*

Four passes are made in order:

| Pass | Action |
|------|--------|
| 1 | Assign directly measured values |
| 2a | Compute group totals (Fines or Coarse) from the sum of their components |
| 2b | Derive a missing component by subtracting the known one from the group total |
| 3 | Apply the 100 % constraint to derive the missing group total |
| 4 | Repeat Pass 2b using group totals that became available in Pass 3 |

This resolves the vast majority of combinations present in the dataset without
any assumptions about the sediment composition.

### Step 3:  Background ratio estimation

For samples that arithmetic derivation could not fully resolve, background ratios are
estimated from all samples where both components of a group are arithmetically known:

$$r_\text{clay/fines} = \text{median}\!\left(\frac{\text{Clay}}{\text{Clay} + \text{Silt}}\right)$$
$$\qquad r_\text{gravel/coarse} = \text{median}\!\left(\frac{\text{Gravel}}{\text{Sand} + \text{Gravel}}\right)$$

A set of overall background proportions (Clay : Silt : Sand : Gravel) is also
derived from samples where all four fractions are known, for use as a final fallback.

### Step 4: Within-group background splits

If a group total (Fines or Coarse) is known but *neither* of its components could be
derived arithmetically, the group total is split using the background ratios from Step 3.

### Step 5: General background imputation

Any fraction still unknown after Steps 2–4 is assigned a share of the remaining
percentage budget (100 % minus the sum of known fractions). The share is proportional
to that fraction's background weight. Fractions already resolved are **never modified**
at this stage.

---

## Quality Confidence Flag

Every sample receives a single `qc_confidence` rating based on accumulated penalty
points from the three pre-processing steps.

```{r}
tibble(
  Source = c(
    "Sum of fractions 100–101 % (minor rounding)",
    "Sum of fractions 101–110 % (moderate conflict)",
    "Sum of fractions 110–200 % (serious conflict)",
    "Sum of fractions > 200 % (likely multiple errors)",
    "Decimal-point correction applied (Step 0b)",
    "Redundant aggregate removed (Step 0c)"
  ),
  `Penalty points` = c("+1", "+2", "+3", "+4", "+1", "+1")
) |>
  kable(align = c("l", "c")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

```{r}
tibble(
  `Total penalty` = c("0", "1", "2", "3", "4 or more"),
  `qc_confidence` = c("high", "medium", "low", "very_low", "unreliable"),
  Interpretation = c(
    "No data quality issues detected",
    "One minor issue (e.g. rounding-only excess or one decimal correction)",
    "Two minor or one moderate issue",
    "Substantial conflict in the original data; values corrected by rescaling",
    "Severe conflict; results may not be reliable"
  )
) |>
  kable(align = c("c", "l", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) |>
  row_spec(5, bold = TRUE, color = "#8B0000")
```

Researchers are encouraged to filter on `qc_confidence` according to the requirements
of their analysis. A common starting point is to retain `"high"` and `"medium"` samples
for quantitative work, and to inspect `"low"` samples on a case-by-case basis.

---

## Derivation Method Labels

In addition to the confidence flag, each of the four output fractions carries a label
describing how it was derived:

```{r}
tibble(
  Method = c("direct", "arithmetic", "background"),
  Meaning = c(
    "The fraction was measured directly in the original data",
    "The fraction was calculated from other measurements using mass-balance arithmetic (no assumptions about composition)",
    "At least one background ratio or overall proportion was required to estimate this fraction"
  )
) |>
  kable() |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE)
```

---

## Data Availability

A tab-delimited files containing particle fractions is available for download at the following location.

 * [pilot_vannmiljo_particles.tsv.gz](https://github.com/seafood-hazards/vannmiljo-pilot/releases/download/v0.1.5/pilot_vannmiljo_particles.tsv.gz)

### Column Definition

The final exported dataset contains one row per sample × sediment layer. The columns
are described below.

```{r}
tibble(
  Column = c(
    "sample_id",
    "sediment_no",
    "clay_pct",
    "silt_pct",
    "sand_pct",
    "gravel_pct",
    "total_pct",
    "clay_method",
    "silt_method",
    "sand_method",
    "gravel_method",
    "any_op_adjusted",
    "qc_confidence"
  ),
  Type = c(
    "character",
    "integer",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "numeric",
    "character",
    "character",
    "character",
    "character",
    "logical",
    "factor"
  ),
  Description = c(
    "Unique sample identifier, linking to the broader dataset",
    "Sediment layer number within the core (1 = topmost measured layer)",
    "Clay fraction (< 2 µm), as a percentage of the total sediment",
    "Silt fraction (2–63 µm), as a percentage",
    "Sand fraction (63–2000 µm), as a percentage",
    "Gravel fraction (> 2000 µm), as a percentage",
    "Sum of the four fractions; equals 100 % for all fully processed samples",
    "Derivation method for clay: 'direct', 'arithmetic', or 'background'",
    "Derivation method for silt",
    "Derivation method for sand",
    "Derivation method for gravel",
    "TRUE if any input measurement had a non-exact operator (i.e. '<', '>', or 'ND')",
    "Data quality confidence level: 'high', 'medium', 'low', 'very_low', or 'unreliable'"
  )
) |>
  kable(align = c("l", "l", "l")) |>
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = TRUE) |>
  column_spec(1, bold = TRUE, monospace = TRUE) |>
  column_spec(2, monospace = TRUE)
```

::: {.callout-note}
Fractions estimated using background ratios (method = `"background"`) are model-derived and carry more
uncertainty than directly measured or arithmetically derived values. The `qc_confidence`
column summarises the overall reliability of each row and should be the primary filter
applied before analysis.
:::

